Supplementary Information

### 2021-01-09 doc/manuscript/MEE_manuscript.Rmd
reference new SI figures

```{r eval=FALSE, echo=FALSE}
## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
#Tables
sites = 1
interactions_and_year_summaries = "S6"

@@

climwin_sameV_sameW_legend = "Figure S5. (PRE at SCBI)"
climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCNM)"

GLS_BCNM_legend = "Figure S9. Best GLS models for Barro Colorado Nature Monument (Panama)"
GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
GLS_SCBI_legend = "Figure S11. Best GLS models for the Smithsonian Conservation Biology Institute (Virginia, USA)"
GLS_LDW_legend = "Figure S12. Best GLS models for Lilley Dickey Woods (Indiana, USA)"
GLS_HF_legend = "Figure S13. Best GLS models for Harvard Forest (Massachusetts, USA) "
GLS_ZOF_legend = "Figure S14. Best GLS models for Zofin Forest (Czech Republic)"
GLS_NE_legend = "Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska, USA)"
GLS_LT_legend = "Figure S16. Best GLS models for Little Tesuque (New Mexico, USA)"
GLS_CB_legend = "Figure S17. Best GLS models for Cedar Breaks (Utah, USA)"
GLS_SC_legend = "Figure S18. Best GLS models for Scotty Creek (Northwest Territory, Canada)"
```

@@

```{r eval=FALSE, echo=FALSE}
## The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:

#Point 1: set the context for and purpose of the work;

#Point 2: indicate the approach and methods;

#Point 3: outline the main results;

#Point 4: identify the conclusions and the wider implications.

# Key-words: A list in alphabetical order not exceeding eight words or short phrases. The most important key-words should appear in the title and the abstract as well as the key-word list. More advice on selecting good keywords can be found here.
```

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. **CREATE AN SI FIGURE!**).

@@

Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested. 
For precipitation variables, interactions were significant for `r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`% of interactions with $RW$ as the growth metric and for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.  `r interactions_fig`).
Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 2 for $BAI$) occurred in species responding positively to precipitation in the current growing season.

Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).

@@

There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_AGB` (depending on growth metric) of the `r n_tested` species- site combinations tested (Fig. `r year_fig`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
In 90% of cases, the growth trend over time was negative.
Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
Significant positive growth trends were observed for only three species, *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%; Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

>>>>

```{r eval=FALSE, echo=FALSE}
## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse, J Matthias

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
#Tables
sites = 1

@@

climwin_sameV_sameW_legend = "Figure S5. (PRE at SCBI)"
climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
climwin_diffV_diffW_legend = "Figure S7. (TMX/TMP at HKK)"

# without year
GLS_BCNM_legend = "Figure S9. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
GLS_BCNM_year_legend = "Figure S10. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
GLS_HKK_legend = "Figure S11. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
GLS_HKK_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
GLS_SCBI_legend = "Figure S13. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
GLS_SCBI_year_legend = "Figure S14. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
GLS_LDW_legend = "Figure S15. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
GLS_LDW_year_legend = "Figure S16. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
GLS_HF_legend = "Figure S17. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
GLS_HF_year_legend = "Figure S18. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
GLS_ZOF_legend = "Figure S19. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
GLS_ZOF_year_legend = "Figure S20. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
GLS_NIO_legend = "Figure S21. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
GLS_NIO_year_legend = "Figure S22. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
GLS_LT_legend = "Figure S23. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
GLS_LT_year_legend = "Figure S24. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
GLS_CB_legend = "Figure S25. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
GLS_CB_year_legend = "Figure S26. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
GLS_SC_legend = "Figure S27. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
GLS_SC_year_legend = "Figure S28. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"

SC_1970break_legend = "Figure 29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."

RW_interactions_all_legend = "Figure S30. (RW_interactions_all)"
BAI_interactions_all_legend ="Figure S31. (BAI_interactions_all)"

@@

```{r eval=FALSE, echo=FALSE}
## Abstract: Limited to 300 words. 
```

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested. 
For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with $RW$ as the growth metric (Fig. `r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with $BAI$ as the growth metric (Fig. `r BAI_interactions_all`).
The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.  `r interactions_fig`).
Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 2 for $BAI$) occurred in species responding positively to precipitation in the current growing season.

Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric (Fig. `r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Fig. `r BAI_interactions_all`).
Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).

@@

There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species- site combinations tested (Figs. `r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; [**ADD SI FIGS--issue 104**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/104)).
In 90% of cases, the growth trend over time was negative.
Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
Significant positive growth trends were observed for only three species, *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
